---
title: "Airlines Fatalities"
author: "Giacomo Maretto"
date: "28 febbraio 2019"
output:
  html_document: default
  pdf_document: default
---


```{r, echo=FALSE,include=FALSE}
#install.packages("rjags","coda","Epi","lme4","pixmap","sp")
library(ggplot2)
library(ggfortify)
library(ggpubr)
library("ggmcmc")
library(rjags)
library(ggplot2)
library(knitr)
library(pander)
library(lattice)
airline <- read.table("http://www.uta.fi/sis/mtt/mttts12/jags/airline.txt",h=T,sep="")
set.seed(123)
```
# Airlines Fatalities

## Introduction 


In this project we are analyzing data collected from 1976 to 2001 about airline fatal accidents by the International Civil Aviation Organization in Montreal , Canada (www.icao.int).

Our goal is to get a good prediction about future fatalitis through a bayesian approach.

Our data is structured with four columns, year, fatal, miles, rate.

"Passenger miles" are in units of $10^{11}$ and the "accident rate" is the number of fatal accidents per $10^{11}$ passenger miles.


```{r, echo=FALSE}
airline$year1975= NULL
pander(head(airline)) # the column fatal which contains the annual number of fatalities.

```

### Following three descriptive time series plot respectively about the number of fatal accidents, the miles flown and the accident rate per each year.


```{r, echo=FALSE}

theme_set(theme_classic())
theme_set(theme_bw())

p = ggplot(airline, aes(x=year)) + 
  geom_line(aes(y=fatal,x =year)) + 
  geom_point(aes(y=fatal,x=year))+
  labs(title="Time Series Chart", 
       subtitle="Returns fatal from 'Airlines' Dataset", 
       caption="Source: American Airlines", 
       y="Fatal")

p  + theme(axis.text.x = element_text(angle = 60, vjust = 0.5)) + scale_x_continuous("Year", labels = as.character(airline$year), breaks = airline$year)


```

```{r, echo=FALSE}

theme_set(theme_classic())
theme_set(theme_bw())

p = ggplot(airline, aes(x=year)) + 
  geom_line(aes(y=miles,x =year)) + 
  geom_point(aes(y=miles,x=year))+
  labs(title="Time Series Chart", 
       subtitle="Returns miles travelled from 'Airlines' Dataset", 
       caption="Source: American Airlines", 
       y="Miles")

p  + theme(axis.text.x = element_text(angle = 60, vjust = 0.5)) + scale_x_continuous("Year", labels = as.character(airline$year), breaks = airline$year)


```

```{r, echo=FALSE}

p = ggplot(airline, aes(x=miles)) + 
  geom_line(aes(y=fatal,x =miles)) + 
  geom_point(aes(y=fatal,x=miles))+
  labs(title="Time Series Chart", 
       subtitle="Returns fatal from 'Airlines' Dataset", 
       caption="Source: American Airlines", 
       y="Fatal")

p  + theme(axis.text.x = element_text(angle = 60, vjust = 0.5)) + scale_x_continuous("Miles", labels = as.character(airline$miles), breaks = airline$miles)


```

```{r, echo=FALSE}
p = ggplot(airline, aes(x=year)) + 
  geom_line(aes(y=miles,x =year)) + 
  geom_point(aes(y=miles,x=year))+
  labs(title="Time Series Chart", 
       subtitle=" Miles per Year", 
       caption="Source: American Airlines", 
       y="Miles")

p  + theme(axis.text.x = element_text(angle = 60, vjust = 0.5)) + scale_x_continuous("Year", labels = as.character(airline$year), breaks = airline$year)
```

```{r, echo=FALSE}
p = ggplot(airline, aes(x=year)) + 
  geom_line(aes(y=rate,x =year)) + 
  geom_point(aes(y=rate,x=year))+
  labs(title="Time Series Chart", 
       subtitle="Rate per Year", 
       caption="Source: American Airlines", 
       y="Rate")

p  + theme(axis.text.x = element_text(angle = 60, vjust = 0.5)) + scale_x_continuous("Year", labels = as.character(airline$year), breaks = airline$year)
```

# JAGS MODELS:

## First model
```{r model 1}
cat( "model {
  for( i in 1:I ) {
    fatal[i] ~ dpois(mu)
    }
    mu ~ dgamma(0.01,0.01)
  }", file="a1.jag" )

a1.par <- c("mu","fatal[27]","fatal[28]","fatal[29]")

a1.ini <- list(list( mu=22 ),
              list( mu=23 ),
              list( mu=24 ) )

a1.dat <- list( fatal = c(airline$fatal,NA,NA,NA), I=29 )

# Model compilation and burn-in
a1.mod <- jags.model( file = "a1.jag",
                    data = a1.dat,
                    inits = a1.ini,
                    n.chains = 3,
                    n.adapt = 1000 )
update(a1.mod,1000)

# Sampling from the posterior
a1.res <- coda.samples( a1.mod,
                      var = a1.par,
                      n.iter = 10000,
                      thin = 10 )
summary( a1.res )

theta <- rgamma(6000, 634, 26 )
y.2002 <- rpois(6000,theta)
plot( main = 'Posterior predictive distribution of y in 2002',table(y.2002), type="h", lwd=5, lend=2, col=gray(0.2), bty="n", ylab="", xlim=c(5,60) )
tpr <- table( as.matrix( a1.res[,"fatal[27]"] ) )
points( as.numeric(names(tpr))+0.4, tpr, type="h", col="red", lwd=4 )
legend("topright", c('Directly simulated','posterior values from JAGS output'),col=c(gray(0.2),'red'),lwd = 5)


```

## Diagnostics for model 1

```{r}

result = ggs(a1.res)
ggs_density(result)
ggs_traceplot(result)
ggs_running(result)
ggs_autocorrelation(result)
```


# Third model

```{r model 3}
library(rjags)
 cat( "model
 {
 for( i in 1:I )
 {
 fatal[i] ~ dpois( mu[i] )
 log(mu[i]) <- alpha + beta*miles[i]+ beta2 * miles[i]*rate[i]
 }
 alpha ~ dnorm(0,0.0001)
 beta ~ dnorm(0.0001,0.00001)
 beta2 ~ dnorm(0.0001,0.00001)
 }",
 
 file="a3.jag" )

 a3.ini= list(
    list(alpha = 0.01, beta = 0.6, beta2=0.1),
    list(alpha = 0.2, beta = 0.2, beta2=0.2),
    list(alpha = 0.7, beta = 0.7, beta2=0.3))
 a3.dat <- list( rate= c(airline$rate,0.7080,0.3,0.4),fatal=c(airline$fatal,NA,NA,NA), miles=c(airline$miles,19.775,23.300,20.3), I=29 )
 a3.par <- c("alpha","beta","beta2","fatal[27]","fatal[28]","fatal[29]")

 # Model compilation and burn-in
 a3.mod <- jags.model(file = "a3.jag",data = a3.dat,inits = a3.ini,n.chains = 3,n.adapt = 1000)
 update(a3.mod,1000)
 # Sampling from the posterior
 a3.res <- coda.samples( a3.mod,var = a3.par,n.iter = 10000,thin = 50 )
 summary( a3.res )

``` 
## Model 3 diagnostics
```{r}
result3 = ggs(a3.res)
ggs_density(result3)
ggs_traceplot(result3)
ggs_running(result3)
ggs_autocorrelation(result3)
```
## Predicting values in model 3
```{r}
par(mfrow=c(1,3))
a3.m <- as.matrix(a3.res)
plot(density(a3.m[,"fatal[27]"]),main="Fatal in 2002", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=14,col="black")
a3.m <- as.matrix(a3.res)
plot(density(a3.m[,"fatal[28]"]),main="Fatal in 2003", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=7,col="black")
a3.m <- as.matrix(a3.res)
plot(density(a3.m[,"fatal[29]"]),main="Fatal in 2004", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=9,col="black")
```
## Forth model

```{r model 4}
library(rjags)
 cat( "model
 {
 for( i in 1:I )
 {
 fatal[i] ~ dpois( mu[i] )
 log(mu[i]) <- alpha + beta2 * miles[i]*rate[i]
 }
 alpha ~ dnorm(0,0.0001)
 
 beta2 ~ dnorm(0.0001,0.00001)
 }",
 
 file="a4.jag" )

 a4.ini= list(
    list(alpha = 0.01, beta2=0.1),
    list(alpha = 0.2, beta2=0.2),
    list(alpha = 0.7,  beta2=0.3))
 a4.dat <- list( rate= c(airline$rate,0.7080,0.3,0.4),fatal=c(airline$fatal,NA,NA,NA), miles=c(airline$miles,19.775,23.300,20.3), I=29 )
 a4.par <- c("alpha","beta2","fatal[27]","fatal[28]","fatal[29]")

 # Model compilation and burn-in
 a4.mod <- jags.model(file = "a4.jag",data = a4.dat,inits = a4.ini,n.chains = 3,n.adapt = 1000)
 update(a4.mod,1000)
 # Sampling from the posterior
 a4.res <- coda.samples( a4.mod,var = a4.par,n.iter = 10000,thin = 50 )
 summary( a4.res )
``` 
## Model 4 diagnostics
```{r}
result4 = ggs(a4.res)
ggs_density(result4)
ggs_traceplot(result4)
ggs_running(result4)
ggs_autocorrelation(result4)
```
## Predicting values in model 4
```{r}
par(mfrow=c(1,3))
a4.m <- as.matrix(a4.res)
plot(density(a4.m[,"fatal[27]"]),main="Fatal in 2002", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=14,col="black")
a4.m <- as.matrix(a4.res)
plot(density(a4.m[,"fatal[28]"]),main="Fatal in 2003", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=7,col="black")
a4.m <- as.matrix(a4.res)
plot(density(a4.m[,"fatal[29]"]),main="Fatal in 2004", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=9,col="black")
```






# Model 5 test with normal approximation

```{r model 5}
library(rjags)
 cat( "model
 {
 for( i in 1:I )
 {
 fatal[i] ~ dnorm( mu[i],1/mu[i] )
 mu[i] <- alpha + beta2 * miles[i]
 }
 alpha ~ dnorm(0,0.0001)
 beta2 ~ dnorm(0,0.0001)
 }",
 file="a5.jag" )

 a5.ini= list(
    list(alpha = 0.01, beta2=3),
    list(alpha = 0.2, beta2=1),
    list(alpha = 0.7, beta2=2))
 a5.dat <- list( rate = c(airline$rate,0.7080,0.3,0.4),fatal=c(airline$fatal,NA,NA,NA), miles=c(airline$miles,19.775,23.300,20.3), I=29 )
 a5.par <- c("alpha","beta2","fatal[27]","fatal[28]","fatal[29]")

 # Model compilation and burn-in
 a5.mod <- jags.model(file = "a5.jag",data = a5.dat,inits = a5.ini,n.chains = 3,n.adapt = 1000 )
 # Sampling from the posterior
 update(a5.mod,1000)
 a5.res <- coda.samples( a4.mod,var = a4.par,n.iter = 3000,thin = 10 )
 summary( a5.res )

``` 

## Model5 diagnostics
```{r}
result5 = ggs(a5.res)
ggs_density(result5)
ggs_traceplot(result5)
ggs_running(result5)
ggs_autocorrelation(result5)
```

## Predicting values in model 5
```{r}
par(mfrow=c(1,3))
a5.m <- as.matrix(a5.res)
plot(density(a5.m[,"fatal[27]"]),main="Fatal in 2002", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=14,col="black")
a5.m <- as.matrix(a5.res)
plot(density(a5.m[,"fatal[28]"]),main="Fatal in 2003", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=7,col="black")
a5.m <- as.matrix(a5.res)
plot(density(a5.m[,"fatal[29]"]),main="Fatal in 2004", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=9,col="black")

```

# Table of true values




      Year      Fatal accidents   Passenger miles    Accident rate
                   

      2002      14                19.775             0.7080
      2003      7                 23.300             0.3004
      2004      9                 20.300             0.4433

## Model comparison throughout DIC index
A good way to select the model that fits best our data, is using the DIC measure.
The DIC (deviance information criterion) is a Bayesian criterion for model comparison, defined as:
$$ DIC = D(\bar{\theta}) + 2_{p_D}$$
 Where: 
      $D(\theta) = -2logL(\theta)$,
      $p_D = \bar D - D(\bar \theta)$, is the effective number of paramenters of the model, the larger $p_D$ the easier the model fits on the data.
      $\bar D = E[D(\theta)]$, its the mean deviance. is a measure of how well the model fits the data, the larger the worse the fit.
      $D(\bar \theta) = D[E(\theta)]$
      
the basic idea is that the models with smaller DIC should be preferred.
Models are penalized both by the value of $\bar D$, which favors a good fit, but also by the effective number of parameters $p_D$.
Since $\bar D$ will decrease as the number of parameters in a model increases, the $p_{D}$ term compensates for this effect by favoring models with a smaller number of parameters.
```{r}
#Model Comparison
DICmodel1 = dic.samples(a1.mod, n.iter=10000, thin=10, type="pD")
DICmodel3 = dic.samples(a3.mod, n.iter=10000, thin =10, type="pD")
DICmodel4 = dic.samples(a4.mod, n.iter=10000, thin=10, type="pD")
DICmodel5 = dic.samples(a5.mod, n.iter=10000, thin =10, type="pD")


DICmodel1
DICmodel3
DICmodel4
DICmodel5
```
# Model 3 is the best fitting one

```{r}
a3.m <- as.matrix(a3.res)

y.pred = c(mean(a3.m[,"fatal[27]"]),mean(a3.m[,"fatal[28]"]),mean(a3.m[,"fatal[29]"]))
y.true = c(14,7,9)
df
df=data.frame(y.true,y.pred)
library(ggplot2)
ggplot( df,aes(x = y.true, y = y.pred)) +  # Set up canvas with outcome variable on y-axis
  geom_point()  # Plot the actual points
```

```{r}
#time series model
miles.new = c(19.775,23.300,20.300)

cat("model {
  trend[1] <- beta0 + beta1 * miles[1]
    for( i in 2 : N ) {
      fatal[i] ~ dnorm( mu[i] , tau)
      mu[i] <- trend[i] + ar1 * ( fatal[i-1] - trend[i-1] )
      trend[i] <- beta0 + beta1 * miles[i]
    }
    ar1 ~ dunif(-1.1,1.1) 
    beta0 ~ dnorm( 0 , 1.0E-12 )
    beta1 ~ dnorm( 0 , 1.0E-12 )
    tau ~ dgamma( 0.001 , 0.001 )}" ,file= "ar1_pred.jag")

ar1_pred.par <- c("ar1","beta0", "beta1", "tau","fatal[27]","fatal[28]","fatal[29]")
ar1_pred.ini <- list( list( ar1=0.01, beta0=0.01, beta1 = 0.01,tau=0.01 ))
ar1_pred.dat <- list( fatal=c(airline$fatal,NA,NA,NA),
                 miles=c(airline$miles,miles.new),
                 N= 26+3)

#mod.forecast = jags.model(ar1_pred.dat,parameters.to.save = ar1_pred.par,mode)

ar1_pred.mod <- jags.model( file = "ar1_pred.jag",
                       data = ar1_pred.dat,
                       inits = ar1_pred.ini,
                       n.chains =1,
                       n.adapt = 1000 )
ar1_pred.res <- coda.samples( model = ar1_pred.mod,
                         var = ar1_pred.par,
                         n.iter = 10000,
                         thin = 10 )
summary(ar1_pred.res)
```
## Model autoregressive diagnostics
```{r}
resultar = ggs(ar1_pred.res)
ggs_density(resultar)
ggs_traceplot(resultar)
ggs_running(resultar)
ggs_autocorrelation(resultar)
```

```{r}
ar.m <- as.matrix(ar1_pred.res)

par(mfrow=c(1,3))
plot(density(ar.m[,"fatal[27]"]),main="Fatal in 2002", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=14,col="black")
plot(density(ar.m[,"fatal[28]"]),main="Fatal in 2003", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=7,col="black")
plot(density(ar.m[,"fatal[29]"]),main="Fatal in 2004", col="red", lwd=2)
grid (NULL,NULL, lty = 6, col = "cornsilk2") 
abline( v=9,col="black")


y.pred = c(mean(ar.m[,"fatal[27]"]),mean(ar.m[,"fatal[28]"]),mean(ar.m[,"fatal[29]"]))
y.true = c(14,7,9)
df
df=data.frame(y.true,y.pred)
library(ggplot2)
ggplot( df,aes(x = y.true, y = y.pred)) +  # Set up canvas with outcome variable on y-axis
  geom_point()  # Plot the actual points
```

